---
title: "6.7 Lab 3: PCR and PLS Regression"
output: 
  github_document:
    md_extensions: -fancy_lists+startnum
  html_notebook: 
    md_extensions: -fancy_lists+startnum
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(pls)
library(ISLR)

# Removing missing values from Hitters dataset
hitters <- ISLR::Hitters %>% na.omit()
```

## 6.7.1 Principal Components Regression

First we apply PCR to predict `Salary`
```{r}
set.seed(2)
pcr_fit <- pcr(Salary ~ .,
  data = hitters, scale = TRUE,
  validation = "CV"
)

summary(pcr_fit)
```

Also one can plot the cross-validation scores (MSE in this case):
```{r}
validationplot(pcr_fit, val.type = "MSEP")
```

The minimum MSE is obtained when we use M = 16 components. However, the CV MSE is roughly equal from M = 1 and upwards, so a model with just one component might suffice.

Now we can perform PCR on the training data and evaluate the test performance:
```{r}
set.seed(1)
hitters_train <- hitters %>% 
  sample_frac(size = 0.5)

hitters_test <- hitters %>% 
  anti_join(hitters_train)

pcr_fit_train <- pcr(Salary ~ ., data = hitters_train, scale = TRUE,
                     validation = "CV")

validationplot(pcr_fit_train, val.type = "MSEP")
```

```{r}
pcr_pred <- hitters_test %>% 
  select(-Salary) %>% 
  predict(pcr_fit_train, ., ncomp = 5)

y_test <- hitters_test %>% pull(Salary)

mean((pcr_pred - y_test)^2)
```

We finally fit PCR on the full data set, using M = 5 (selected from the PCR on training data):
```{r}
pcr_fit_full <- pcr(Salary ~ ., data = hitters, scale = TRUE, ncomp = 5)
summary(pcr_fit_full)
```

## 6.7.2 Partial Least Squares

```{r}
set.seed(1)
pls_fit <- plsr(Salary ~ ., data = hitters_train,
                scale = TRUE, validation = "CV")

summary(pls_fit)
```

In this case, the lowest CV error ocurrs with M = 1. Now we can evaluate the test MSE:
```{r}
pls_pred <- hitters_test %>% 
  select(-Salary) %>% 
  predict(pls_fit, ., ncomp = 1)

mean((pls_pred - y_test)^2)
```

Finally we perform PLS on the full data, with M = 1.
```{r}
pls_fit_full <- plsr(Salary ~ ., data = hitters, scale = TRUE, ncomp = 1)
summary(pls_fit_full)
```

