---
title: "8.3 Lab: Decision Trees"
output: 
  github_document:
    md_extensions: -fancy_lists+startnum
  html_notebook: 
    md_extensions: -fancy_lists+startnum
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(tree)
library(ISLR)
library(modelr)
library(MASS) # for Boston dataset
library(randomForest)
library(gbm)
```

## 8.3.1 Fitting Classification Trees

We first analyze the `Carseats` data set. We first encode the continuous variable `Sales` as a binary one.
```{r}
( 
carseats <- ISLR::Carseats %>% 
  as_tibble() %>% 
  mutate(High = factor(ifelse(Sales <= 8, "No", "Yes")))
)
```

Now we try to predict `High` using all variables but `Sales`.
```{r}
tree_carseats <- tree(High ~ . - Sales, data = carseats)
```


The `summary()` function lists the variables that are used as internal nodes in the tree, the number of terminal nodes, and the (training) error rate.
```{r}
summary(tree_carseats)
```

Displaying the tree structure:
```{r, fig.asp=1, out.width=5}
plot(tree_carseats)
text(tree_carseats, pretty = 0)
```

If we just type the name of the tree object, R prints output corresponding to each branch of the tree. R displays:

* the split criterion (e.g. `Price<92.5`)
* the number of observations in that branch
* the deviance, the overall prediction for the branch (Yes or No)
* the fraction of observations in that branch that take on values of Yes and No

Branches that lead to terminal nodes are indicated using asterisks.

```{r}
tree_carseats
```

Obtaining the test error: 
```{r}
set.seed(1989)

carseats_train <- carseats %>% 
  sample_frac(0.5)

carseats_test <- carseats %>% 
  anti_join(carseats_train)
```


```{r}
tree_carseats <- tree(High ~ . - Sales, data = carseats_train)

carseats_test <- carseats_test %>% 
  add_predictions(tree_carseats, type = "class")

caret::confusionMatrix(data = carseats_test$pred,
                       reference = carseats_test$High)
```

Now let's consider whether pruning the tree might lead to improved results.

```{r}
set.seed(1989)
cv_carseats <- cv.tree(
  tree_carseats,
  FUN = prune.misclass
)
names(cv_carseats)
```

```{r}
cv_carseats
```

`$dev` represents the CV classification error. We can plot it against the number of nodes:
```{r}
qplot(cv_carseats$size, cv_carseats$dev, geom = "line") +
  geom_vline(xintercept = cv_carseats$size[which.min(cv_carseats$dev)],
             color = "red")
```

In this case, the tree with less CV error has 12 nodes. We can obtain this pruned tree using `prune.misclass()`
```{r, fig.asp=0.8, out.width=5}
prune_carseats <- prune.misclass(tree_carseats, best = 12)
plot(prune_carseats)
text(prune_carseats, pretty = 0)
```

How does the pruned tree perform on the test data set?
```{r}
carseats_test_pred_pruned <- 
  carseats_test %>% 
  add_predictions(prune_carseats, type = "class")

caret::confusionMatrix(
  data = carseats_test_pred_pruned$pred,
  reference = carseats_test_pred_pruned$High
)
```

Accuracy went up from 71% to 76.5%.

## 8.3.2 Fitting Regression Trees

```{r}
set.seed(1989)

boston_train <- Boston %>% 
  as_tibble() %>% 
  sample_frac(0.5)

boston_test <- Boston %>% 
  as_tibble() %>% 
  anti_join(boston_train)
```

```{r}
tree_boston <- tree(medv ~ . , data = boston_train)

summary(tree_boston)
```

Here "deviance" is just the sum of the squared errors (RSS)

```{r}
plot(tree_boston)
text(tree_boston, pretty = 0)
```

Checking if pruning the tree improves the performance:
```{r}
cv_boston <- cv.tree(tree_boston)

qplot(cv_boston$size, cv_boston$dev, geom = "line") +
  geom_vline(xintercept = cv_boston$size[which.min(cv_boston$dev)],
             color = "red")

```

The tree selected by CV has 7 terminal nodes.
```{r}
prune_boston <- prune.tree(tree_boston, best = 7)

plot(prune_boston)
text(prune_boston, pretty = 0)
```

The tree lost the node that splits by `dis < 2.845`.

Now we can do prediction using the pruned tree, and see how the predicted values relate to the actual values of `medv`.
```{r}
boston_test <- boston_test %>% 
  add_predictions(prune_boston)

qplot(pred, medv, data = boston_test)
```

Measuring the test set MSE
```{r}
(
  test_mse_boston <- 
  mean((boston_test$pred - boston_test$medv)^2)
)
```

```{r}
sqrt(test_mse_boston)
```

This means that the model leads to test predictions that are within around $4898 of the true median home value for the suburb.

## 8.3.3 Bagging and Random Forests

We now use the `randomForest` package. Note that bagging it's just a special case of random forest where $m = p$, so the same package can be used for bagging.

Performing bagging:
```{r}
set.seed(1989)

bag_boston <- randomForest(medv ~ ., data = boston_train,
                           mtry = 13, importance = TRUE)

bag_boston
```

`mtry` indicates the number of predictors that should be considered at each split (in this case, all the available predictors, since we're doing bagging).

Now let's measure the test error:
```{r}
boston_test <- boston_test %>% 
  add_predictions(bag_boston, var = "pred_bag")

qplot(pred_bag, medv, data = boston_test)
```

```{r}
(
  test_mse_bag <- 
  mean((boston_test$pred_bag - boston_test$medv)^2)
)
```

```{r}
sqrt(test_mse_bag)
```

The test MSE is much lower using bagging than using a single decision tree. Let's see if the test error further decreases when using random forest instead.

```{r}
set.seed(1989)

rf_boston <- randomForest(medv ~ ., data = boston_train,
                          mtry = 6, importance = TRUE)

rf_boston
```

```{r}
boston_test <- boston_test %>% 
  add_predictions(rf_boston, var = "pred_rf")

(
  test_mse_rf <- 
  mean((boston_test$pred_rf - boston_test$medv)^2)
)
```

The test MSE in fact decreases from 12.55 to 11.

We can use the `importance()` function to see the importance of each variable:
```{r}
importance(rf_boston)
```

Two measures of variable importance are reported. The former is based upon the mean decrease of accuracy in predictions on the out of bag samples when a given variable is excluded from the model. The latter is a measure of the total decrease in node impurity that results from splits over that variable, averaged over all trees.

Plotting variable importance ranking:
```{r}
varImpPlot(rf_boston)
```

## 8.3.4 Boosting

Finally we use the `gbm::gbm()` function to fit boosted regression trees.

To fit boosted regression trees we need to specify `distribution = "gaussian"`. To use classification trees instead, we use `distribution = "bernoulli`.
```{r}
set.seed(1989)

boost_boston <- gbm(medv ~ .,
                    data = boston_train,
                    distribution = "gaussian",
                    n.trees = 5000,
                    interaction.depth = 4)

summary(boost_boston)
```

The function `summary()` provides a variable importance table and plot. We see that `lstat` and `rm` are the most important variables by far.

We can produce partial dependence plots for certain variables, i.e. illustrate the marginal effect of a given variable after "integrating out" the other variables.

```{r}
plot(boost_boston, i = "rm")
plot(boost_boston, i = "lstat")
```

Using the model to predict on the test set:
```{r}
predictions_boost <- 
  predict(boost_boston, new_data = boston_test, n.trees = 5000, type = "response")

mean((predictions_boost - boston_test$medv)^2)
```

For some reason, the test MSE is much higher than in boosting and random forest. Let's try with other tuning parameters:
```{r}
boost_boston2 <- 
  gbm(medv ∼ .,
      data = boston_train, distribution = "gaussian", n.trees =5000 , interaction.depth = 4,
      shrinkage =0.2, verbose =F)

predictions_boost2 <- 
  predict(boost_boston2, new_data = boston_test, n.trees = 5000, type = "response")

mean((predictions_boost2 - boston_test$medv)^2)

```

